---
title: "Decoding Delays: Analyzing Toronto’s Streetcar System"
subtitle: "Investigating Patterns and Trends in TTC Streetcar Delays (2022-2024)"
author: 
  - Daniel Du
thanks: "Code and data are available at https://github.com/danield424/TTC-Streetcar-Delays"
date: September 24, 2024
abstract: "This paper analyzes delay data from Toronto's TTC streetcar system between 2022 and 2024 to assess patterns in delay frequency and length. The study found that while the frequency of delays has decreased, the average delay length has increased, with delays showing no consistent patterns based on month, day of the week, or hour. Understanding streetcar delay trends helps determine methods of improving public transportation in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

library(tidyverse)
library(knitr)
library(dplyr)
library(lubridate)
library(patchwork)
library(ggplot2)
```

# Introduction

  Public transportation systems are vital components of urban infrastructure, providing essential services that connect people to their communities, reduce traffic congestion, and promote sustainability. Streetcars were once a common mode of transportation across North American cities, but have largely disappeared due to car-centric urban planning @britannica. Today, Toronto stands out as the only city on the continent still utilizing a streetcar network, with the Toronto Transit Commission (TTC) running 11 daytime and 5 nighttime lines. These streetcars are often criticized for being slow and unreliable, with frequent delays frustrating commuters while contributing to broader traffic congestion, undermining the positives of public transit @thestar. Addressing these inefficiencies could significantly improve accessibility, boost sustainability, and enhance the overall quality of life for in Canada’s largest city.

  Despite the central role of the streetcar, which serves approximately 230,000 daily commuters, there is a lack of comprehensive analysis quantifying the impact of streetcar delays on transit efficiency @cp24. This paper explores the frequency, duration, and patterns of these disruptions and aims to highlight patterns. Understanding where, when, and why these delays occur is central to identifying the areas most in need of improvement and optimizing this key component of Toronto’s public transportation system.

  This study analyzes public streetcar delay data from 2022 to 2024 to assess the extent and impact of delays in Toronto's streetcar network. After examining delay patterns across different years, months, days, and hours, and comparing average delay lengths and frequencies across various service lines, the analysis reveals a uniform consistent increase in average delay times, despite a reduction in delay frequency. These findings highlight systemic inefficiencies in the streetcar network, showing the need for further research and investment to improve transit performance. The findings from this study support movements to enhance the efficiency and reliability of Toronto’s public transportation system.

  The paper is structured as follows: @sec-data introduces the data and methodology used; @sec-results presents the key findings; and @sec-discussion discusses the real-world implications of these results. @sec-limitations discusses the limitations of the dataset and potential avenues for future research.

# Data {#sec-data}

  To investigate streetcar delays, delay logs from 2022, 2023, and 2024 (up until September) is sourced from the @opendatatoronto R package, which provides data from the City of Toronto Open Data Catalogue. These datasets, intended to monitor and improve public transit performance, are regularly updated and were current as of September 1, 2024. Although similar datasets exist, this specific dataset was chosen due to its official nature, reliability, and size.

  The data includes the date and time, line number, and delay length of over 40,000 instances of streetcar delays since 2022. It is unclear how the data was gathered; reliability is ensured as it comes directly from the TTC, but it is possible there were missed or incorrect measurements due to limitations. Delay times are measured in minutes, and time is measured to the second.

  The data was thoroughly cleaned to ensure accuracy and prepare it for analysis; the process is detailed in Appendix~\ref{sec:appendixA}. In @sec-2-1 and @sec-2-2 we explore the dataset variables before conducting further analysis in @sec-results.

## Overview of Data {#sec-2-1}

@fig-1 provides a sample of 6 data observations. In total, there are 36701 observations of 5 variables: `day` contains the day of the week, `date` contains the timestamp of the delay, `service_type` was constructed and assigns the service type of the line in `line`. `min_delay` is the delay time in minutes; this is the main variable of interest. We will analyze `min_delay` in relation to the 4 other variables. There are no additional relationships between the variables aside from the obvious `day` - `date` and aforementioned `service_type` construction.

```{r}
#| label: fig-1
#| fig-cap: Summary of Streetcar Delay Data 2022-2024
#| echo: false

# Read data and display a summary with head().
streetcar_data <- read_csv("../data/analysis_data/delaydata.csv", show_col_types = FALSE)
kable(head(streetcar_data), 
  format = "pipe")
```

## Delay by Year and Service Type  {#sec-2-2}

The number of delay observations per year and for each service type are below. Note the low number of delays in 2024 is due to the data ending in August. The vast majority of observations come from the regular streetcar service, which is intuitive since the regular service has the most lines (9 total) and runs the most regularly.

```{r}
#| label: fig-2
#| echo: false
# Table with number of delay observations by year.
delays_by_year <- streetcar_data %>%
  mutate(year = year(date)) %>% 
  count(year)                   
kable(delays_by_year, 
  caption = "Number of Delay Observations by Year")

# Table with the number of delay observations by service type
delays_by_service_type <- streetcar_data %>%
  count(service_type)
kable(delays_by_service_type, 
  caption = "Number of Delay Observations by Service Type")
```


We can calculate summary statistics for the delay length, and then filter by year and service type:

\newpage



```{r}
#| label: fig-3
#| echo: false
#| results: asis

# Extract summary statistics
min_delay_summary <- summary(streetcar_data$min_delay)
min_delay_table <- data.frame(
  Statistic = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max."),
  Value = as.numeric(min_delay_summary[c(1, 2, 3, 4, 5, 6)]))

# Display the formatted summary table
kable(min_delay_table,
  booktabs = TRUE,
)

# Summary by Service Type
service_type_summary <- streetcar_data %>%
  group_by(service_type) %>%
  summarize(
    mean_delay = round(mean(min_delay), 1),
    median_delay = median(min_delay),
    sd_delay = round(sd(min_delay), 1),
    minimum_delay = min(min_delay),
    maximum_delay = max(min_delay)
  )

kable(service_type_summary,
      booktabs = TRUE,
  caption = "Min. Delay Summary by Service Type")


year_summary <- streetcar_data %>%
  group_by(year = year(date)) %>%
  summarize(
    mean_delay = round(mean(min_delay), 1),
    median_delay = median(min_delay),
    sd_delay = round(sd(min_delay), 1),
    minimum_delay = min(min_delay),
    maximum_delay = max(min_delay)
  )

kable(year_summary,
      booktabs = TRUE,
  caption = "Min Delay Summary by Year")

```

The minimum delay of 3 minutes and maximum of 119 minutes are due to the constraints we placed when cleaning the data, as discussed in Appendix~\ref{sec:appendixA}. We can see that the median delay is twice as long for the night service compared to the two others. The median delay times over the 3 years are similar, although the mean delay rose by more than 3 minutes between 2022 and 2024.

Following the introduction of the variables, we can explore the results of our data analysis.

# Results {#sec-results}

## Analysis of Average Delay Time

One of the key factors of interest is the average delay time for different variable values, such as streetcar lines, types, and dates. Examining average delays allows us to identify patterns and areas where delays are most pronounced. This highlights specific points of inefficiency and potential targets for improvement. In @fig-avg1 we begin by looking at the average delay time for each line.

```{r}
#| label: fig-avg1
#| fig-cap: Average delay time for streetcar lines, sorted by service type.
#| echo: false

# Calculate the average delay for each line 
avg_delay_data <- streetcar_data %>%
  group_by(line, service_type) %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop')

# Reorder service_type to put "Regular" first and "Night" last
avg_delay_data$service_type <- factor(avg_delay_data$service_type, levels = c("Regular", "Reduced", "Night"))

# Bar graph it
ggplot(avg_delay_data, aes(x = as.factor(line), y = avg_delay, fill = service_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Delay Time for Streetcar Lines",
    x = "Streetcar Line", y = "Average Delay Time (mins)") +
  scale_fill_manual(values = c("Regular" = "red3", "Reduced" = "tomato", "Night" = "blue4")) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "none") +
  facet_grid(~ service_type, scales = "free_x", space = "free")
```

Average delay times vary significantly across streetcar lines and service types. Regular service lines have relatively consistent delay times, with lines 503 and 505 experiencing slightly higher averages than others. Night service lines, particularly line 305, exhibit the highest average delays. Reduced service lines also show higher delay times compared to regular service lines. In @fig-avg2 we see if this is the case for each year.

```{r}
#| label: fig-avg2
#| fig-cap: Average delay time for streetcars by service type, per year.
#| echo: false

# Calculate the average delay for each service type and year
avg_delay_by_year <- streetcar_data %>%
  mutate(year = year(date)) %>%  
  group_by(service_type, year) %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop')

# Reorder service_type to put "Regular" first and "Night" last
avg_delay_by_year$service_type <- factor(avg_delay_by_year$service_type, levels = c("Regular", "Reduced", "Night"))
# Create a new variable that combines year and service type for unique colours
avg_delay_by_year$year_service <- interaction(avg_delay_by_year$year, avg_delay_by_year$service_type)

ggplot(avg_delay_by_year, aes(x = service_type, y = avg_delay, fill = year_service)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Delay Time by Service Type and Year",
    x = NULL, y = "Average Delay Time (mins)") +
  scale_fill_manual(
    values = c(
      "2022.Regular" = "red1", "2022.Reduced" = "tomato1", "2022.Night" = "steelblue2",
      "2023.Regular" = "red3", "2023.Reduced" = "tomato3", "2023.Night" = "steelblue3",
      "2024.Regular" = "red4", "2024.Reduced" = "tomato4", "2024.Night" = "blue4"
    ),  
    name = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right")
```

Delay times have increased over the years across all service types, with the most significant increase seen in reduced service lines, particularly from 2022-2023. In years 2023 and 2024, both reduced and night services had a significantly higher average delay length. We can now look at specific lines for each type individually, starting with regular service in @fig-avg3 and @fig-avg4.

```{r}
#| label: fig-avg3
#| fig-cap: Average delay time for regular service streetcars, per year.
#| echo: false
 
# Filter data for regular service, and calculate the average delay for each line and year
regular_data <- streetcar_data %>%
  filter(service_type == "Regular") %>%
  mutate(year = year(date)) %>% 
  group_by(line, year)

avg_delay_regular <- regular_data %>% summarize(avg_delay = mean(min_delay), .groups = 'drop')

# Bar graph with 3 bars for each line per year.
ggplot(avg_delay_regular, aes(x = as.factor(line), y = avg_delay, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Delay for Regular Service Lines by Year",
    x = "Streetcar Line", y = "Average Delay Time (mins)") +
  scale_fill_manual(
    values = c("2022" = "red1", "2023" = "red3", "2024" = "red4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right")
```

Across almost all regular lines, average delay times have increased over time, indicating a worsening in service reliability. Line 503 and 511 are the regular lines with the most concerning delay time and upward trend, while 512 is the only line with a downward trend.

```{r}
#| label: fig-avg4
#| fig-cap: Average delay time for regular service streetcars over time.
#| echo: false

# Get the month_year to plot as x-axis, and calculate average delay over time
reg_delay_over_time <- regular_data %>%
  mutate(month_year = as.Date(format(date, "%Y-%m-01"))) %>%
  mutate(month_year = case_when(month_year == as.Date("2024-09-01") ~ as.Date("2024-08-01"),
      TRUE ~ month_year)
  ) %>%
  group_by(line, month_year) %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop')

# Calculate the overall average delay for each month-year across all lines
overall_reg_delay <- reg_delay_over_time %>%
  group_by(month_year) %>%
  summarize(overall_reg_delay = mean(avg_delay))

# Merge overall average with the filtered data
avg_delay_with_overall <- reg_delay_over_time %>%
  left_join(overall_reg_delay, by = "month_year")

# Line graph showing average delay over time for each regular streetcar line
ggplot(avg_delay_with_overall, aes(x = month_year, y = avg_delay, colour = as.factor(line))) +
  geom_line(linewidth = 0.6, alpha = 0.6) +
  geom_line(aes(y = overall_reg_delay), linetype = "dashed", color = "black", linewidth = 0.8) +  # Overall average delay line
  labs(
    title = "Delay Time of Regular Service Streetcar Lines",
    x = NULL, y = "Average Delay Time (mins)",
    colour = "Streetcar Line") +
  scale_x_date(limits = c(as.Date("2022-01-01"), as.Date("2024-09-01")),
               breaks = seq(as.Date("2022-01-01"), as.Date("2024-09-01"), by = "3 months"),
              labels = c("Jan '22", "", "Jul '22", "", "Jan '23", "", "Jul '23", "", "Jan '24", "", "Jul '24")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

In @fig-avg4 we can examine each streetcar line's growth individually over time. Growth is not linear and many lines share similar patterns of delay increases and decreases, particularly during common periods. For instance, peaks and troughs in delays across several lines around mid-2023 suggest external factors impacting multiple lines simultaneously. 

Most lines align with the overall average delay trend, dotted in black. Line 503's high peaks and variance stand out compared to the rest, suggesting that it has unique factors or circumstances influencing its delays. The 512, previously shown to be the only line with a decreasing average delay time over the years, appears to have had a temporary period of delay times under 10 minutes starting in late 2023, before reverting back to being near the average.

Similarly, @fig-avg5, @fig-avg6, and @fig-avg7 analyze reduced and night services over time.

```{r}
#| label: fig-avg5
#| fig-cap: Average delay time for reduced and night service streetcars, per year.
#| echo: false

# Same process as above code chunk
# Reduced
reduced_data <- streetcar_data %>%
  filter(service_type == "Reduced") %>%
  mutate(year = year(date)) %>% 
  group_by(line, year)

avg_delay_reduced <- reduced_data %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop') %>%
  # Ensure all combinations of line and year are present
  complete(line = c(507, 508), year = c(2022, 2023, 2024), fill = list(avg_delay = 0))


reduced_plot <- ggplot(avg_delay_reduced, aes(x = as.factor(line), y = avg_delay, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Reduced Service Avg. Delay",
    x = NULL, y = "Average Delay Time (mins)") +
  scale_fill_manual(
    values = c("2022" = "tomato1", "2023" = "tomato3", "2024" = "tomato4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    legend.position = "right")

# Night
night_data <- streetcar_data %>%
  filter(service_type == "Night") %>%
  mutate(year = year(date)) %>% 
  group_by(line, year)

avg_delay_night <- night_data %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop') %>%
  # Ensure all combinations of line and year are present
  complete(line = c(301, 304, 305, 306, 310), year = c(2022, 2023, 2024), fill = list(avg_delay = 0))


night_plot <- ggplot(avg_delay_night, aes(x = as.factor(line), y = avg_delay, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Night Service Avg. Delay",
    x = NULL, y = NULL) +
  scale_fill_manual(
    values = c("2022" = "steelblue2", "2023" = "steelblue3", "2024" = "blue4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    legend.position = "right")

print(reduced_plot + night_plot + plot_layout(ncol = 2))
```

We can now see that there are no observations of the reduced service 507 in 2022 as it was yet to be routed. Therefore, the increase in average delay time for the reduced lines can be attributed entirely to the 508. Similarly, the night service 305 was not routed until 2024, and has by far the lengthiest average delays.


```{r}
#| label: fig-avg6
#| fig-cap: Average delay time for reduced service streetcars over time.
#| echo: false

# Same as avg-delay-4
red_delay_over_time <- reduced_data %>%
  mutate(month_year = as.Date(format(date, "%Y-%m-01"))) %>%
  mutate(month_year = case_when(month_year == as.Date("2024-09-01") ~ as.Date("2024-08-01"),
      TRUE ~ month_year)
  ) %>%
  group_by(line, month_year) %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop')

overall_red_delay <- red_delay_over_time %>%
  group_by(month_year) %>%
  summarize(overall_red_delay = mean(avg_delay))

red_delay_with_overall <- red_delay_over_time %>%
  left_join(overall_red_delay, by = "month_year")

ggplot(red_delay_with_overall, aes(x = month_year, y = avg_delay, colour = as.factor(line))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_line(aes(y = overall_red_delay), linetype = "dashed", color = "black", linewidth = 0.6) + 
  labs(
    title = "Delay Time of Reduced Service Streetcar Lines",
    x = NULL, y = "Average Delay Time (mins)",
    colour = "Streetcar Line") +
  scale_x_date(limits = c(as.Date("2022-01-01"), as.Date("2024-09-01")),
               breaks = seq(as.Date("2022-01-01"), as.Date("2024-09-01"), by = "3 months"),
              labels = c("Jan '22", "", "Jul '22", "", "Jan '23", "", "Jul '23", "", "Jan '24", "", "Jul '24")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

Since the 507 line was in service, the average delay time for the two reduced lines follow similar trends, indicating a correlation between the two. Both peaked late in 2023 and have decreased since.

```{r}
#| label: fig-avg7
#| fig-cap: Average delay time for night service streetcars over time.
#| echo: false

# Same as avg-delay-4
n_delay_over_time <- night_data %>%
  mutate(month_year = as.Date(format(date, "%Y-%m-01"))) %>%
  mutate(month_year = case_when(month_year == as.Date("2024-09-01") ~ as.Date("2024-08-01"),
      TRUE ~ month_year)
  ) %>%
  group_by(line, month_year) %>%
  summarize(avg_delay = mean(min_delay), .groups = 'drop')

overall_n_delay <- n_delay_over_time %>%
  group_by(month_year) %>%
  summarize(overall_n_delay = mean(avg_delay))

n_delay_with_overall <- n_delay_over_time %>%
  left_join(overall_n_delay, by = "month_year")

ggplot(n_delay_with_overall, aes(x = month_year, y = avg_delay, colour = as.factor(line))) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_line(aes(y = overall_n_delay), linetype = "dashed", color = "black", linewidth = 0.6) + 
  labs(
    title = "Delay Time of Night Service Streetcar Lines",
    x = NULL, y = "Average Delay Time (mins)",
    colour = "Streetcar Line") +
  scale_x_date(limits = c(as.Date("2022-01-01"), as.Date("2024-09-01")),
               breaks = seq(as.Date("2022-01-01"), as.Date("2024-09-01"), by = "3 months"),
              labels = c("Jan '22", "", "Jul '22", "", "Jan '23", "", "Jul '23", "", "Jan '24", "", "Jul '24")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

There does not appear to be much of a correlation between night service lines, with most of them appearing to peak randomly. The overall average delay time is often skewed by lengthy delays of over 40 minutes. The average delay time has plenty of variance for each line, except the 301 which appears fairly consistent. 

While analysis of the average delay time is useful and helps identify general trends and areas of growth, looking at the frequency of delays gives us a clearer picture of trends and line efficiency.

## Analysis of Delay Frequency

Before we proceed with delay frequency, we have to deal with missing values in 2024 from September onwards, which we do in Appendix~\ref{sec:appendixB}. We do this by only using data from January to August when making year-by-year comparisons. Since the total observations for each service type are different, we will look at each service type individually.

```{r}
#| label: fig-freq1
#| fig-cap: Total delays in January-August for regular service lines by year.
#| echo: false

# Keep only January to August
delay_freq_regular1to8 <- regular_data %>%
  filter(month(date) %in% 1:8) %>% summarize(delay_freq = n(), .groups = 'drop')

# Calculate the average delay frequency for each line and year within the filtered period
ggplot(delay_freq_regular1to8, aes(x = as.factor(line), y = delay_freq, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Delay Frequency for Regular Service Lines (Jan-Aug)",
    x = NULL, y = "Total Delays") +
  scale_fill_manual(
    values = c("2022" = "red1", "2023" = "red3", "2024" = "red4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    legend.position = "right")
```

@fig-freq1 shows the delay frequency by line from January to August of the 3 years. Since the dataset does not provide information on successful trips, a high delay frequency can either mean operational issues or a high service volume, so looking at the value of a frequency alone is not insightful. However, we can see an overall downward trend in delay frequency over the years.


```{r}
#| label: fig-freq2
#| fig-cap: Total delays in January-August for reduced and night service lines by year.
#| echo: false

# Same code as above
# Reduced
delay_freq_reduced1to8 <- reduced_data %>%
  filter(month(date) %in% 1:8) %>% summarize(delay_freq = n(), .groups = 'drop') %>% 
  # Ensure all combinations of line and year are present
  complete(line = c(507, 508), year = c(2022, 2023, 2024), fill = list(delay_freq = 0))

reduced_plot2 <- ggplot(delay_freq_reduced1to8, aes(x = as.factor(line), y = delay_freq, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Reduced Service Delays (Jan-Aug)",
    x = NULL, y = "Total Delays") +
  scale_fill_manual(
    values = c("2022" = "tomato1", "2023" = "tomato3", "2024" = "tomato4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    legend.position = "right")

# Night
delay_freq_night1to8 <- night_data %>%
  filter(month(date) %in% 1:8) %>% summarize(delay_freq = n(), .groups = 'drop') %>% 
  complete(line = c(301, 304, 305, 306, 310), year = c(2022, 2023, 2024), fill = list(delay_freq = 0))

night_plot2 <- ggplot(delay_freq_night1to8, aes(x = as.factor(line), y = delay_freq, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Night Service Delays (Jan-Aug)",
    x = NULL, y = NULL) +
  scale_fill_manual(
    values = c("2022" = "steelblue2", "2023" = "steelblue3", "2024" = "blue4"),
    name = NULL) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    legend.position = "right")

print(reduced_plot2 + night_plot2 + plot_layout(ncol = 2))
```

Not many conclusions can be drawn from the reduced service graph given the sample size. The night service graph has a bit more information, with overall delays decreasing significantly from 2022 to 2023 before increasing again in 2024. The tiny sample size for the 305 indicates that its high average delay in @fig-avg5 may not necessarily be an indicator of poor performance, but rather that more data is required.

Up until now, we have mainly looked at patterns in streetcar performance over time, but there are also elements to be found in researching more specific time frames.

## Specific Time Periods

For further data exploration, we look at the hour, day of the week, month, and year. Details for acquiring this data are in Appendix~\ref{sec:appendixC}. Since the sample size for reduced and night service lines are much smaller than regular service lines, little information is gained from analyzing their specific time periods, so we will look at regular service data.

```{r}
#| echo: false
# Extract hour, day of the week, month for regular streetcar rides.
reg_time_data <- streetcar_data %>% filter(service_type == "Regular") %>%
  mutate(hour = hour(date), 
    day_of_week = wday(date, label = TRUE),
    month = factor(format(date, "%b"), levels = month.abb),
    year = year(date))
reg_time_data <- reg_time_data %>% filter(!(reg_time_data$month == 'Sep' & reg_time_data$year == 2024))
```

@fig-time1 and @fig-time2 visualize the average delay time and delay frequency for each month.
@fig-time3 and @fig-time4 explore the same for each day of the week.
@fig-time5, @fig-time6, and @fig-time7 analyze the hourly breakdown.

```{r}
#| label: fig-time1
#| fig-cap: Total number of regular service delays by month.
#| echo: false

ggplot(reg_time_data, aes(x = as.factor(month), fill = as.factor(year))) +
    geom_bar() +
    labs(
      title = "Regular Service Delays by Month",
      x = NULL, y = "# of Delays",
      fill = NULL) +
    scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

```

```{r}
#| label: fig-time2
#| fig-cap: Average length of regular service delays by month.
#| echo: false

# Plot average delay time by month
ggplot(reg_time_data, aes(x = month, y = min_delay, fill = as.factor(year))) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge") +
  labs(
    title = "Average Delay Time for Regular Service by Month",
    x = NULL, 
    y = "Average Delay Time (Minutes)",
    fill = NULL
  ) +
  scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

The number of service delays seem to be lowest in the summer and highest in the winter, which suggests the weather may have a large affect on streetcar punctuality. The average delay time appears to be lowest at the start of the year and gradually increases as the year goes on. 

```{r}
#| label: fig-time3
#| fig-cap: Total number of service delays by day of week.
#| echo: false

# Plot for day of week
ggplot(reg_time_data, aes(x = day_of_week, fill = as.factor(year))) +
    geom_bar() +
    labs(title = "Regular Service Delays by Day of Week",
      x = NULL, y = "# of Delays",
      fill = NULL) +
    scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

```{r}
#| label: fig-time4
#| fig-cap: Average length of regular service delays by day of week.
#| echo: false

ggplot(reg_time_data, aes(x = day_of_week, y = min_delay, fill = as.factor(year))) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge") +
  labs(
    title = "Average Delay Time for Regular Service by Day of Week",
    x = "Day of Week", 
    y = "Average Delay Time (Minutes)",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

```

In 2022, the number of service delays were lowest on Sunday and increased as the week went along, but this trend has disappeared; 2024 had roughly a uniform number of delays for each day of the week. The same is true for the average delay time, which seems to have no correlation with the day of the week.

```{r}
#| label: fig-time5
#| fig-cap: Total number of regular service delays by hour of the day.
#| echo: false

# Plot for hour of the day
ggplot(reg_time_data, aes(x = as.factor(hour), fill = as.factor(year))) +
    geom_bar() +
    labs(
      title = "Regular Service Delays by Hour of the Day",
      x = "Hour of Day", y = "# of Delays",
      fill = NULL) +
    scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

```


```{r}
#| label: fig-time6
#| fig-cap: Total number of night service delays by hour of the day.
#| echo: false

night_time_data <- streetcar_data %>% filter(service_type == "Night") %>%
  mutate(hour = hour(date), 
    day_of_week = wday(date, label = TRUE),
    month = factor(format(date, "%b"), levels = month.abb),
    year = year(date)) 

night_time_data <- night_time_data %>% filter(!(night_time_data$month == 'Sep' 
                                                & night_time_data$year == 2024)) %>% complete(hour = 0:23, 
                                                                                              fill = list(count = 0))
ggplot(night_time_data, aes(x = as.factor(hour), fill = as.factor(year))) +
    geom_bar() +
    labs(
      title = "Night Service Delays by Hour of the Day",
      x = "Hour of Day", y = "# of Delays",
      fill = NULL) +
    scale_fill_manual(values = c("2022" = "steelblue2", "2023" = "steelblue3", "2024" = "blue4")) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), 
      legend.position = "none")

```

The number of delays for the regular service dips between 4 and 10 am, but this seems to entirely be because of reduced service volume at those hours. The night service delay total corroborates this, as the total peaks at the same time the regular service dips. The delay peak around 6-9pm for regular service coincides with a rush hour, but this peak lessens year on year from 2022 to 2024.

```{r}
#| label: fig-time7
#| fig-cap: Average length of regular service delays by hour.
#| echo: false

ggplot(reg_time_data, aes(x = as.factor(hour), y = min_delay, fill = as.factor(year))) +
  stat_summary(fun = "mean", geom = "bar", position = "dodge") +
  labs(
    title = "Average Delay Time for Regular Service by Hour",
    x = "Hour of Day", 
    y = "Average Delay Time (Minutes)",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("2022" = "red1", "2023" = "red3", "2024" = "red4")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

The average delay time has a clear peak around 6 to 8 am. Aside from this peak, delay time is uniform throughout.

# Discussion {#sec-discussion}

## Trend of Increasing Average Delay Lengths

  The analysis shows a general upward trend in average delay times each year across all service types and specific time intervals. This consistent increase could be the result of rowing traffic congestion or operational issues The exact reason for this increase is unclear; it could be from systemic issues within the streetcar network, infrastructure inefficiencies, or just a rise in population. Whatever the case, addressing these delays should be paramount; if there is no simple solution, there are ways to mitigate the problem, such as enhancing real-time communication with commuters through online platforms like Twitter or the TTCwatch app to manage expectations.

## Decreasing Frequency of Delays

  There is a general trend of the frequency of delays decreasing over the observed period. However, this decrease could be misleading. Since the dataset does not have total service volume, it is unclear if the reduced frequency of delays is due to operational improvements or reduced service levels. If fewer streetcars are operating due to maintenance, route cuts, or lack of demand, the lower frequency may not represent an actual improvement in streetcar reliability. Further analysis with complete service data would be helpful.


## Independence of Delay Patterns from Specific Times

  Delay lengths and frequencies appear to be mostly independent of the month, day of the week, or hour of the day, suggesting that there are no outlier areas where delays are a big issue, and any issue with delays cannot be stopped easily with targeted short-term fixes. This could mean that delays are driven by broader systemic factors; addressing these will require even more comprehensive research and long-term strategizing.

# Limitations and Future Steps {#sec-limitations}

- Lack of data on total trip count: The absence of data on the total number of streetcar trips for each line makes it challenging to determine whether variance in delay frequency is due to actual changes in delay patterns, or changes in the total volume of services. Future research should incorporate data on total streetcar trips to provide a more accurate context and allow for more robust conclusions about service reliability.

- Missing future data: The absence of data from September to December 2024 limits the ability to fully assess seasonal trends, particularly during the autumn months, which may impact overall year-to-year comparisons. As time goes on, more detailed data will be available, increasing our understanding of delays and whether existing trends will continue to hold.

- Lack of location data: Not having data on the specific locations of lines and stops limits the ability to explore spatial patterns and relationships between different streetcar lines, which would increase knowledge on the interactions between lines and service types. Collecting more detailed data, including line coordinates, intersections, and distances would provide a clearer picture of how delays are related.


\newpage

\appendix

# Appendix {-}

\section{Data Cleaning}\label{sec:appendixA}
This included constructing a variable to determine the type of service - regular, reduced, night - of each specific streetcar line, and filtering observations to only include streetcar lines currently in service @ttc. Additionally, delays under 2 minutes were removed and considered negligible, as were delays over 2 hours, as these were likely to be announced stoppages. Other information in the dataset such as station location, incident type, bound, and vehicle number were not discussed in this analysis. Data was cleaned using @citeR and packages @tidyverse, @dplyr, and @janitor. Data was analyzed with @ggplot2, @lubridate, @tidyverse, @dplyr, @patchwork, and @knitr.


\section{Incomplete Data}\label{sec:appendixB}
One possible consideration would be to scale the 2024 data to account for a whole year's worth of observations. However, this is an unsafe estimate as it assumes delays are evenly distributed throughout the year, which is unlikely considering seasonal changes, unexpected occurrences, and planned events like construction. For best accuracy, when we directly compare between years, we will use data from January to August, which is present in all 3 years of the dataset.

\section{Time Data}\label{sec:appendixC}
We get the year, month, and hour from the `date` variable, and use the existing `day` variable to assign an ordinal variable `day_of_week`.

# References

\newpage
